---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sanssouci

<!-- badges: start -->
  [![R build status](https://github.com/pneuvial/sanssouci/workflows/R-CMD-check/badge.svg)](https://github.com/pneuvial/sanssouci/actions)
[![Coverage Status](https://codecov.io/gh/pneuvial/sanssouci/branch/develop/graph/badge.svg)](https://codecov.io/github/pneuvial/sanssouci/branch/develop)
<!-- badges: end -->

The goal of sanssouci [sãsusi] is to perform **post hoc inference**: in a multiple testing context, sanssouci provides statistical guarantees on possibly user-defined and/or data-driven sets of hypotheses. 

Typical use cases include:

- **Differential gene expression (DGE) studies in genomics**: 
  - see vignettes dedicated to [microarray data](https://pneuvial.github.io/sanssouci/articles/post-hoc_differential-expression.html) and [RNAseq data](https://pneuvial.github.io/sanssouci/articles/post-hoc_differential-expression_RNAseq.html) 
  - run [volcano plot shiny app](https://shiny-iidea-sanssouci.apps.math.cnrs.fr/)
- **fMRI studies in neuroimaging**: see [fMRI vignette](https://pneuvial.github.io/sanssouci/articles/post-hoc_fMRI.html). 

In both cases, the permutation-based post hoc inference methods implemented in the `sanssouci` package outperform classical post hoc bounds based on probabilistic inequalities.

## Example: differential analysis

We briefly illustrate the case of differential gene expression study (see dedicated [vignette](https://pneuvial.github.io/sanssouci/articles/post-hoc_differential-expression.html) for more information). We start by creating an object of class `sanssouci` from gene expression data available from the [sanssouci.data](https://github.com/pneuvial/sanssouci.data) package, 

```{r load-data}
library("sanssouci")
#> remotes::install_github("pneuvial/sanssouci.data")
data(expr_ALL, package = "sanssouci.data")
groups <- ifelse(colnames(expr_ALL) == "BCR/ABL", 1, 0) # map to 0/1
obj <- SansSouci(Y = expr_ALL, groups = groups)
```

Then we fit the method by with `B=1000` permutations, and with a target risk `alpha = 0.1`.

```{r fit, cache=TRUE}
res <- fit(obj, alpha = 0.1, B = 1000)
```

### Output 1 - Post hoc bound for a subset of genes

Let us assume that we are interested in genes with a $p$-value below $10^{-3}$. 

```{r user-selection}
S <- which(pValues(res) < 1e-3)
predict(res, S)
```

The method ensures with $1-\alpha = 90\%$ confidence that there are at least `r predict(res, S=S, what = "TP")` truly differentially expressed genes (true positives, TP) among these  `r length(S)` genes, corresponding to a false discovery proportion (FDP) less than `r ceiling(predict(res, S=S, what = "FDP")*100)/100`.


### Output 2 - Confidence curves for "top-k" feature lists

Another output of the method is the following $1-\alpha$-level confidence curve on FDP (left panel) and corresponding TP (right panel) in "top-k" gene lists. 

```{r conf-curve}
library("ggplot2")
plot(res, xmax = 500) +
  geom_vline(xintercept = length(S), 
             color = "gray", linetype = "dotted", size = 1.5) +
  geom_line(size = 1.5)
```


### Output 3 - Volcano plots

A more refined user selection is obtained by selecting genes based on both significance (here, a FDR-adjusted $p$-value below 0.05) and fold change (here, fold change above 0.3). Such selections are generally represented by volcano plots:

```{r volcano-plot}
volcanoPlot(res, q = 0.05, r = 0.3, ylim = c(0, 6))
```

Importantly, multiple such selections can be made without compromising the validity of the bounds. The [IIDEA shiny app](https://shiny-iidea-sanssouci.apps.math.cnrs.fr/) makes it possible to perform such selections interactively and without programming. 

## Installation

<!-- You can install the released version of sanssouci from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("sanssouci") -->
<!-- ``` -->

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("pneuvial/sanssouci")
```

