---
title: "Joint Error Rate calibration"
subtitle: "Simulations for one and two-sample tests"
author: "P. Neuvial"
date: "2018-03-27"
output: 
  html_document: 
    toc: yes
vignette: >
  %\VignetteIndexEntry{Joint Error Rate calibration}
  %\VignetteEngine{knitr::knitr}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This vignettes illustrates the following points

* simulation of one- and two-sample Gaussian equi-correlated observations
* computation of test statistics by randomization
* calibration of Joint-Family-Wise Error Rate (JER) thresholds

We start with two-sample tests because we believe they are used more frequently.

```{r library}
library("sanssouci")
#set.seed(0xBEEF) # for reproducibility
library("ggplot2")
```

Parameters:

```{r params}
m <- 5e2
n <- 30
pi0 <- 0.8
rho <- 0.3
SNR <- 3
```

We use the function `SansSouciSim` to generate Gaussian equi-correlated samples. We then 'fit' the resulting SansSouci object

## Two-sample tests

### Simulation

```{r simulation}
obj <- SansSouciSim(m, rho, n, pi0, SNR = SNR, prob = 0.5)
obj
```

We perform JER calibration using the linear (Simes) template $\mathfrak{R(\lambda)}=(R_1(\lambda),$ $\ldots,R_K(\lambda)$, where for $1 \leq k \leq K$

$$R_k(\lambda) = \left\{i\in \{1, \dots , m\}\::\: \bar{\Phi}(Z_i) >  \frac{\lambda k}{m} \right\}\,$$

where $\bar{\Phi}$ is the cdf of the $\mathcal{N}(0,1)$ distribution. Note that $p_i = \bar{\Phi}(Z_i)$ is the one-sided $p$-value associated to the test statistics $Z_i$.

### Calibration

```{r calibration}
B <- 1000
alpha <- 0.2
cal <- fit(obj, alpha = alpha, B = B, family = "Simes")
cal
```

The output of the calibration is the following `SansSouci` object. In particular:

- `cal$input` contains the input simulated data (and the associated truth)
- `cal$param` contains the calibration parameters
- `cal$output` contains the output of the calibration, including
   -  `p.values`: $m$ test statistics calculated by permutation
   - `thr` : A JER-controlling family of $K$ (here $K=m$) elements
   - `lambda`: the $\lambda$-calibration parameter

Because we are under positive equi-correlation, we expect $\lambda > \alpha$ for the Simes family. 

```{r cal-Simes}
cal$output$lambda
cal$output$lambda > alpha
```

### Post hoc confidence bounds

The fitted `SansSouci` object contains post hoc confidence bounds:

```{r confidence-bounds}
env <- predict(cal, all = TRUE)
head(env)
```

We compare it to the true number of false positives among the most significant items, and to the Simes bound without calibration 

```{r other-conf-env}
cal0 <- fit(obj, alpha = alpha, B = 0, family = "Simes")
oracle <- fit(obj, alpha = alpha, family = "Oracle")

confs <- list(Simes = predict(cal0, all = TRUE),
              "Simes+calibration" = predict(cal, all = TRUE),
              "Oracle" = predict(oracle, all = TRUE))
```

```{r plot-conf-env}
plotConfCurve(confs, xmax = 200)
```


## One sample tests

The code is identical, except for the line to generate the observations (where we do not specify a probability of belonging to one of the two populations using the `prob` argument); moreover it is not necessary to specify a vector of categories 'categ' in 'calibrateJER'.

```{r, sim-one-sample}
obj <- SansSouciSim(m, rho, n, pi0, SNR = SNR)
obj
```

```{r cal-one-sample}
cal <- fit(obj, alpha = alpha, B = B, family = "Simes")
cal
```

Again we expect $\lambda > \alpha$.

### Confidence curves

The associated confidence curves are displayed below:


```{r other-conf-env-one-sample}
cal0 <- fit(obj, alpha = alpha, B = 0, family = "Simes")
oracle <- fit(obj, alpha = alpha, family = "Oracle")

confs <- list(Simes = predict(cal0, all = TRUE),
              "Simes+calibration" = predict(cal, all = TRUE),
              "Oracle" = predict(oracle, all = TRUE))
```

```{r plot-conf-env-one-sample}
plotConfCurve(confs)
```

## Session information

```{r session-info}
sessionInfo()
```

