---
title: "Permutation-based post hoc inference for fMRI studies"
author: "Alexandre Blain and Pierre Neuvial"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
  pdf_document: default
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 1
bibliography: sansSouci.bib
vignette: |
  %\VignetteIndexEntry{Permutation-based post hoc inference for fMRI studies} 
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{sansSouci.data}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
fig.width=8,
fig.height=6
)
```


This vignette shows how to use the [`sansSouci`](https://github.com/pneuvial/sanssouci) package to perform permutation-based post hoc inference for fMRI data. We focus on a motor task (left versus right click) from the Localizer data set by @orfanos2017brainomics. The data set is included in the [`sansSouci.data`](https://github.com/pneuvial/sanssouci.data) package. It has been obtained using the Python module [nilearn](http://nilearn.github.io/), see @abraham2014machine-learning.

# Setup

We first make sure that the required R packages are installed and loaded.

```{r install-r-packages, results='hide', message=FALSE, eval=FALSE}
require("ggplot2") || install.packages("ggplot2")
require("sansSouci") || remotes::install_github("pneuvial/sanssouci@develop")
require("sansSouci.data") || remotes::install_github("pneuvial/sanssouci.data")
```


```{r load-r-packages, results='hide', message=FALSE}
library("ggplot2")
library("sansSouci")
```

We set the seed of the random number generator for numerical reproducibility of the results:
```{r set-seed}
set.seed(20200924)
```


# Loading and preprocessing the data

```{r load-data}
data(fMRI_localizer, package = "sansSouci.data")
dat <- fMRI_localizer
rm(fMRI_localizer)
```

The samples are classified into two subgoups, These subgroups correspond to two different motor tasks the subjects were asked to perform, "left click" vs "right click":

```{r colnames}
m <- nrow(dat)
categ <- colnames(dat)
knitr::kable(table(categ))
```

# Post hoc inference by the Simes inequality

We start by calculating the Simes bound introduced by @GS2011. This bound has already been applied to fMRI data, see @rosenblatt2018ari. In that paper, post hoc inference (by the Simes inequality) is referred to as "All-resolutions inference" (ARI).

```{r set-alpha}
alpha <- 0.1
```

We set the target confidence level to $\alpha = `r alpha`$. The Simes bound may be obtained with the function 'curveSimes' of the R package 'cherry'. Here, we use the 'sansSouci' package, as follows:


```{r simes-bound}
categ <- ifelse(colnames(dat) == "left", 0, 1) # map to 0/1
dex <- data.frame(rowWelchTests(dat, categ))
rwt <- rowWelchTests(dat, categ=categ)
pvals <- rwt$p.value
ce_Simes <- confidenceEnvelope(pvals, refFamily = "Simes", param = alpha)
```

Disregarding (for now) the voxel locations in the brain, we plot confidence bounds on $p$-value level sets (corresponding to maps obtained by thresholding at a given test statistic):

```{r plot-conf-bounds-Simes}
xmax <- 8*max(subset(ce_Simes, stat == "TP")$bound)
plotConfidenceEnvelope(ce_Simes, xmax = xmax) +
  ggtitle("Post hoc confidence envelopes: Simes/ARI")
```

# Tighter confidence bounds by adaptation to unknown dependence 

As discussed in @blanchard20post-hoc, the above-described bound is known to be valid only under certain positive dependence assumptions (PRDS) on the joint $p$-value distribution. Although the PRDS assumption is widely accepted for fMRI studies (see @genovese2002thresholding, @nichols2003controlling), we argue (and demonstrate below) that this assumption yields overly conservative post hoc bounds. Indeed, the Simes bound is by construction not *adaptive* to the specific type of dependence at hand for a particular data set.


To bypass these limitations, @blanchard20post-hoc have proposed a randomization-based procedure known as $\lambda$-calibration, which yields tighter bounds that are adapted to the dependency observed in the data set at hand. We note that a related approach has been proposed by @HSG2019. In the case of two-sample tests, this calibration can be achieved by permutation of class labels, which is readily available via the `calibrateJER` function of the sansSouci package:


```{r post-hoc-calibration, cache=TRUE}
B <- 1000
cal <- calibrateJER(X = dat, categ, B = B, alpha = alpha, refFamily = "Simes")
```

An alternative to the Simes/Linear reference family is the Beta reference family:

```{r post-hoc-calibration-beta, cache=TRUE}
K <- 500
cal_beta <- calibrateJER(X = dat, categ, B = B, alpha = alpha, refFamily = "Beta", K = K)
```


```{r plot-conf-bounds}
conf_envs <- list("Simes/ARI" = ce_Simes, 
                "Linear"= cal$conf_env,
                "Beta"= cal_beta$conf_env)
names(conf_envs)[3] <- sprintf("Beta (K=%s)", K)


cols <- RColorBrewer::brewer.pal(length(conf_envs), "Dark2")
p <- plotConfidenceEnvelope(conf_envs, xmax = xmax, cols = cols)
p + geom_line(size=1.5)
```

In this example, ARI is outperformed by permutation-based approaches, which are able to adapt to unknown dependency: for the same target risk level ($\alpha = `r alpha`$), both permutation-based bounds are much less conservative than the classical Simes/ARI bound. 

# Post hoc bounds on brain atlas areas

The goal in this section is to calculate post hoc bound on user-defined brain regions. One definition of such regions is given by the [Harvard-Oxford brain atlas](https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/Atlases), which is available from the sansSouci.data package. 

```{r load-atlas}
data(brainAtlas, package = "sansSouci.data")
```

This atlas is made of `r nrow(brainAtlas)` areas. For each of these areas, we will calculate the three post hoc bounds obtained above. Note that by construction of post hoc bounds, these 48 bounds are valid simultaneously (in particular, no further multiple testing correction is required at this stage). 

```{r thr}
thrs <- list("Simes/ARI" = SimesThresholdFamily(nrow(dat))(alpha),
             "Linear" = cal$thr,
             "Beta" = cal_beta$thr)
names(thrs)[3] <- sprintf("Beta (K=%s)", K)
```

For convenience we define a function to calculate these post hoc bounds for all areas from the chosen atlas:

```{r get-posthoc-bounds}
getPosthocBounds <- function(reg, pvals, thrList) {
  nreg <- nrow(reg)
  FP <- matrix(NA, ncol = length(thrList), nrow = nreg)
  colnames(FP) <- names(thrList)
  R <- numeric(nreg)
  for (rr in 1:nreg) {
    wpos <- which(reg[rr, ]>0)
    pvalsRR <- pvals[wpos]  ## collapse all into one big region (good idea?)
    for (kk in seq(along = thrList)) {
      FP[rr, kk] <- maxFP(pvalsRR, thrList[[kk]])
    }
    R[rr] <- length(pvalsRR)
  }
  data.frame(area = rownames(reg), size = R, R - FP, check.names = FALSE)
}
```


```{r posthoc-bounds-atlas}
bounds <- getPosthocBounds(brainAtlas, pvals, thrs)
```

In this particular case, some signal is revealed by the post hoc bound for `r sum(rowSums(bounds[, 4:5])>0)` areas for permutation-based approaches:

```{r posthoc-bounds-atlas-table}
ww <- which(rowSums(bounds[, 4:5])>0)
cap <- sprintf("Lower bounds on the number of true positives in %s brain regions.",
               length(ww))
knitr::kable(bounds[ww, ], caption = cap)
```

Here again, ARI is outperformed by permutation-based approaches which are able to adapt to unknown dependency: for the same target risk level ($\alpha = `r alpha`$), both permutation-based bounds are much less conservative than the classical Simes/ARI bound.

# Session information

```{r session-info}
sessionInfo()
```

# Reproducibility

To re-build this vignette from its source, use: 

```{r reproducibility, eval = FALSE}
rmarkdown::render("post-hoc_fMRI.Rmd", output_format = "pdf_document")
# To keep intermediate files, add option 'clean = FALSE'
rmarkdown::render("post-hoc_fMRI.Rmd", output_format = "html_document")
```



# References