---
title: "Conservativeness of the Simes inequality under positive dependence"
author: "P. Neuvial"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
header-includes:
  - \newcommand{\cH}{}
  - \renewcommand{\P}{\mathbb{P}}
bibliography: sansSouci.bib
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Conservativeness of the Simes inequality under positive dependence}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Simes inequality

Let $\mathbf{q}=(q_i)_{1 \leq i \leq m}$ be a vector of random variables such that:

* for each $i=1 \ldots m$, $q_i \sim \mathcal{U}[0,1]$
* $\mathbf{q}$ satisfies positive regression dependency on a subset (PRDS)

We refer to @sarkar98probability for a formal definition of this form of positive dependence. Here, we simply note that it holds in particular of Gaussian equi-correlated random variables. 

Then, letting $(q_{(1)}, \ldots q_{(m)} )$ be the corresponding order statistics, we have:

$$ \mathbb{P} \left( \exists i \in \{1, \ldots m\} \::\: q_{(i)} \leq \frac{\alpha i}{m}\right) \leq \alpha $$
This inequality is due to @simes86improved. It is sharp under independence of the $q_i$ (that is, the above inequality is an equality). How sharp is it under positive dependence?

## Estimating the size of the Simes test

```{r library}
library("sansSouci")
library("matrixStats")
library("knitr")
```

We write a small function to estimate the size of the Simes test, that is, the level actually achieved by the left-hand side in the above inequality.

```{r size-Simes}
size <- function(rho, m, B, alpha=0.05) {
    X0 <- gaussianTestStatistics(m, B, dep = "equi", param = rho)$X0
    p0 <- pnorm(X0, lower.tail = FALSE)
    sk <- sansSouci::SimesThresholdFamily(m, kMax = m)
    sansSouci:::empiricalCoverage(sk(alpha), X0)
}
```

Our simulation parameters are set as follows:

```{r params}
m <- 1e3
rhos <- c(0, 0.1, 0.2, 0.4, 0.8)
alpha <- 0.2
B <- 10000
```

That is:

* $`r m`$ hypotheses
* equi-correlation level between $`r rhos[1]`$ and $`r rhos[length(rhos)]`$
* target level of the Simes test: $`r alpha`$
* $`r B`$ replications are used to estimate the size of the test


```{r simulations}
ahat <- sapply(rhos, size, m, B, alpha=alpha)
ses <- sqrt(ahat*(1-ahat)/B)
```

We estimate the size of the test as the average size $\hat{\alpha}$ achieved across replications, and  the corresponding standard error as $\sqrt{\hat{\alpha}(1-\hat{\alpha})/B}$.

```{r results-table, results="as.is"}
tb <- rbind(ahat/alpha, ses/alpha)
colnames(tb) <- rhos
rownames(tb) <- c("Achieved level/target level", "Standard error")
knitr::kable(tb, digits = 2)
```

This table illustrates the sharpness of the Simes test under independence ($\rho=0$), and the conservativeness of this test under positive dependence ($\rho>0$). For example, when $\rho=0.4$, the size od the Simes test is less than half the target level.

## Session information

```{r}
sessionInfo()
```

## References