---
title: "Step-down joint Family-Wise Error Rate control"
author: "Pierre Neuvial"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Step-down joint Family-Wise Error Rate control}
---

This is an internal vignette which illustrates the notion of Joint Family-Wise Error Rate (JER) control, and the possible benefits of *step-down* control over single step JER control.  For simplicity (and historical reasons) we chose here to directly simulate the test statistics (and not the underlying observations). The JER calibration is perfomed by assuming known joint distribution of the test statistics under the null hypothesis.

```{r}
library("sansSouci")
```

## Simulation settings

Parameters:

```{r}
m <- 2e2
n <- 123
pi0 <- 0.5
B <- 1e3
rho <- 0.2
```

We use the function `gaussianTestStatistics` to generate Gaussian equi-correlated test statistics directly (without passing by observations and associated tests). We then use the (internal) function `calibrateJER0` to perform the calibration.

```{r}
#set.seed(0xBEEF)
sim <- gaussianTestStatistics(m, B, pi0 = pi0, SNR = 3, dep = "equi", param = rho)
X0 <- sim$X0
x <- sim$x
H0 <- which(sim$H == 0)
H1 <- which(sim$H == 1)
```

## Test statistics

```{r, testStat, fig.width=7, fig.height=5}
pch <- 20
plot(x, col = 1 + sim$H, main = "Test statistics", pch=pch)
legend("topleft", c("H0", "H1"), pch = pch, col = 1:2)
```

## Step-down JER control

```{r}
alpha <- 0.1
res <- sansSouci:::calibrateJER0(X0, refFamily = "kFWER", alpha=alpha, stat=x)
thr <- res$thr
thrMat <- res$stepsDown$thr
```

## Confidence envelopes
```{r}
o <- order(x, decreasing=TRUE)
xo <- x[o]
bounds <- apply(thrMat, 2, function(thr) curveMaxFP(xo, thr))
```

## True number of false discoveries among first rejections
```{r}
V <- cumsum(o %in% H0)
```

## "Oracle" JER thresholds
```{r}
X0.Oracle <- X0[-H1, ]
x.Oracle <- x[-H1]
res.Oracle <- sansSouci:::calibrateJER0(X0.Oracle, refFamily = "kFWER", alpha = alpha, stat = x.Oracle)
thrO <- c(res.Oracle$thr, rep(-Inf, length(H1)))
xo.Oracle <- x.Oracle[order(x.Oracle, decreasing = TRUE)]
VbarO <- curveMaxFP(xo, thrO)
```

## Graphical illustration
```{r, JER_SD, fig.width=7, fig.height=5}
nSteps <- ncol(thrMat)
cols1 <- seq.int(nSteps)
ltys1 <- rep(1, nSteps)
ttag <- paste("m=", m, ", rho=", rho, ", alpha=", alpha, sep = "")
ttl <- paste("Bounds on #FP among rejected hypotheses", ttag, sep = "\n")
xmax <- min(200, m)
ymax <- bounds[xmax, 1]
matplot(bounds, t = 's', lty = 1, ylab = "V", col = cols1,
        main = ttl, xlim = c(1, xmax), ylim = c(0, ymax))

cols2 <- c("purple", "pink")
ltys2 <- rep(1, 2)
lines(V, col = cols2[1], t = "s", lty = ltys2[1])
lines(VbarO, col = cols2[2], t = "s", lty = ltys2[2])

lgd <- c(paste("SD-JER(step=", 1:nSteps, ")", sep = ""), "True V", "Oracle JER")
ltys <- c(ltys1, ltys2)
cols <- c(cols1, cols2)
legend("top", lgd, col = cols, lty = ltys)

xSc <- x/max(x)*ymax  ## scaled test statistics (for display)
colStat <- 1 + sim$H
points(xSc[o], col = colStat[o], pch = pch, cex = 0.2)
legend("left", c("H0", "H1"), pch = pch, col = 1:2)
```

## Session information

```{r}
sessionInfo()
```

